---
title: "ST5209/X Assignment 1"
format: pdf
editor: visual
author: "Wang Zixuan,A0298286H"
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Clone the course [repo](https://github.com/yanshuotan/st5209-2024).
3.  Create a separate folder in the root directory of the repo, label it with your name, e.g. `yanshuo-assignments`
4.  Copy the assignment1.qmd file over to this directory.
5.  Modify the duplicated document with your solutions, writing all R code as code chunks.
6.  When running code, make sure your working directory is set to be the folder with your assignment .qmd file, e.g. `yanshuo-assignments`. This is to ensure that all file paths are valid.[^1]

[^1]: You may view and set the working directory using `getwd()` and `setwd()`.

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
```

## Question 1 (Quarto)

Read the [guide](https://quarto.org/docs/computations/r.html) on using Quarto with R and answer the following questions:

a)  Write a code chunk that imports `tidyverse` and `fpp3`.

```{r}
library(tidyverse)
library(fpp3)
```

a)  Modify the chunk so that only the following output is shown (i.e. the usual output about attaching packages and conflicts is not shown.)

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(fpp3)
```

echo=FALSE: Hides the **code** but shows the **output**.

message=FALSE: Suppresses package loading **messages**.

warning=FALSE: Hides **warnings**.

\includegraphics[width=0.9\textwidth]{../_images/quarto_chunk_output.png}

c)  Modify the chunk so that it is executed but no code is shown at all when rendered to a pdf.

```{r setup, include=FALSE}
library(tidyverse)
library(fpp3)
```

Use the *include=FALSE* option. So the **code and output** will be completely hidden in the rendered PDF, but the code **still runs** in the background.

c)  Modify the document so that your name is printed on it beneath the title.

Modify the YAML section in the beginning of the document.

## Question 2 (Livestock)

Consider the `aus_livestock` dataset loaded in the `fpp3` package.

a)  Use `filter()` to extract a time series comprising the monthly total number of pigs slaughtered in Victoria, Australia, from Jul 1972 to Dec 2018.

```{r}
#view(aus_livestock)
aus_livestock |>   
  filter(Animal=="Pigs",
         State=="Victoria",
         Month >= yearmonth("1972 Jul")& Month <= yearmonth("2018 Dec"))
```

a)  Make a time plot of the resulting time series.

```{r}
aus_livestock |>   
  filter(Animal=="Pigs",
         State=="Victoria",
         Month >= yearmonth("1972 Jul")& Month <= yearmonth("2018 Dec")) |>
  autoplot(.vars = Count)+
  labs(
    title = "Monthly Total Number of Pigs Slaughtered in Victoria, Australia",
    x = "Year",) +
  theme_minimal()
```

## Question 3 (Beer production)

Consider the `aus_production` dataset loaded in the `fpp3` package. We will study the column measuring the production of beer.

a)  Make a time plot of the beer production time series.

```{r}
#view(aus_production )
aus_production|>
  autoplot(.vars = Beer)+
  labs(
    title = "Beer Production in Australia",
    x = "Year",
    y = "Beer Production") +
  theme_minimal()
```

a)  Describe the observed trend.

We see that the time series displays seasonality. Also, the beer production was increasing before 1975. After 1975, the beer production became stable, but still maintained a seasonal pattern.

a)  Make a seasonal plot.

```{r}
aus_production|>
  gg_season(Beer)+
  labs(
    title = "Seasonal Plot of Beer Production in Australia") +
  theme_minimal()
```

a)  What is the period of the seasonality?

We see that the time series displays the **4 quarter** seasonality(i.e. 1 year). The beer production is higher in Q4 and Q1, and decreases in Q2, thus the beer production is lower in Q2 and Q3.

a)  Describe the seasonal behavior.

Since Australia's summer falls in **Q4 and Q1**, which includes Christmas, New Year. People want to celebrate the occasions, thus beer consumption **increase**. On the contrary, as temperatures drop, beer consumption **declines** in **Q2 and Q3** since people tend to prefer warm beverages.

## Question 4 (Pelts)

Consider the `pelt` dataset loaded in the `fpp3` package, which measures the Hudson Bay Company trading records for Snowshoe Hare and Canadian Lynx furs from 1845 to 1935.

a)  Plot both time series on the same axes. *Hint: Use `pivot_longer()` to create a key column*.

```{r}
#view(pelt)
pelt_long <- pelt |>  
  pivot_longer(cols = c(Hare, Lynx), 
               names_to = "Species", values_to = "Furs") #a new key column

pelt_long |>  
  autoplot(Furs) +  
  labs(
    title = "Hudson Bay Company Trading Records",
    x = "Year",
    y = "Number of Furs Traded"
  ) +
  theme_minimal()
```

a)  What happens when you try to use `gg_season()` to the lynx fur time series? What is producing the error?

```{r}
#pelt_long |> 
 # gg_season(Lynx)
```

*We get error: "! The data must contain at least one observation per seasonal period."*

The error occurs because gg_season() expects a regular time series. But this dataset only contains yearly index. So there is no within-year seasonality to plot.

a)  Make a lag plot with the first 20 lags. Which lags display strong positive correlation? Which lags display strong negative correlation? Verify this with the time plot.

```{r}
plt1 <- pelt |> gg_lag(Lynx, lags = 1:20, geom = "point", alpha = 0.2)+
  labs(title = "Lag Plot of Lynx Furs")
plt2 <- pelt |> gg_lag(Hare, lags = 1:20, geom = "point", alpha = 0.2)+
  labs(title = "Lag Plot of Hare Furs")
library(gridExtra)
grid.arrange(plt1, plt2, ncol = 2)
```

For the lag plots for the Lynx, The graphs with lags of 1, 10, and 20 months show strong positive correlations. Lag plots of 5 and 15 months show strong negative correlations.

For the lag plots for the Hare, The graphs with lags of 1, 10 months show strong positive correlations. Lag plots of 5 and 15 months show strong negative correlations.

a)  If you were to guess the seasonality period based on the lag plot, what would it be?

There may be a cycle pattern. Base on the lag plot, the seasonality period could be around 10 years.

a)  Use the provided function `gg_custom_season()` in `_code/plot_util.R`[^2] to make a seasonal plot for lynx furs with the period that you guessed.[^3] Does the resulting plot suggest seasonality? Why or why not?

[^2]: You can load this function using `source("../_code/plot.util.R")`.

[^3]: Unfortunately, it seems \``gg_season()` does not allow this functionality.

```{r}
#| warning: FALSE
source("../_code/plot_util.R")
pelt |>
  gg_custom_season(Lynx, period = 10)
pelt |>
  gg_custom_season(Hare, period = 10)
```

From the plot, we can see a 10-year period seanonality. The data increases in the first 5 years and decreases in the second 5 years. However, the seasonality is not that conspicuous. So maybe there is also a cycle in this time series.

## Question 5 (Box-Cox, Q3.3 in FPP)

Why is the Box-Cox transform unhelpful for the `canadian_gas` data?

```{r}
var(canadian_gas$Volume)
canadian_gas |> 
  autoplot(Volume)
```

 

```{r}
lambda <- canadian_gas |>
  features(Volume, features = guerrero) |>
  pull(lambda_guerrero)
canadian_gas |>
  autoplot(box_cox(Volume, lambda))
```

The Box-Cox transform is used to stabilize variance. However, the fluctuations throughout the time series do not have similar variability to the level of the series. So for this 1-parameter transformations, the Box-Cox transform is unhelpful.

## Question 6 (Decomposition with outliers, Q3.7 in FPP)

Consider the last five years of the Gas data from `aus_production` .

```{r}
gas <- tail(aus_production, 5*4) |> select(Gas)
#view(gas)
```

a.  Plot the time series. Can you identify seasonal fluctuations and/or a trend-cycle?

    ```{r}
    gas |> autoplot() +
      labs(title = "Gas Production",
           x = "Year", y = "Gas Production") +
      theme_minimal()
    ```

    We can observe a yearly seasonality and increasing trend.

b.  Use `classical_decomposition` with `type=multiplicative` to calculate the trend-cycle and seasonal indices.

    ```{r}
    #| warning: false
    gas_decomp <- gas |>
      model(classical_decomposition(Gas, type = "multiplicative"))|>
      components()

     gas_decomp |>
        autoplot()+
      labs(title = "Classical Decomposition of Gas Production",
           x = "Year", y = "Gas Production")+
         theme_minimal()
    ```

    By using the time series decomposition, the trend is upward trend. And the time series has a 1-year seasonality. The gas production is higher in Q2 and Q3, lower in Q1 and Q4.

c.  Do the results support the graphical interpretation from part a?

    Yes, they come to the same interpretation.

d.  Compute and plot the seasonally adjusted data.

    ```{r}
    library(ggplot2)
    ggplot(gas_decomp, aes(x = Quarter, y = season_adjust)) +
      geom_line(color = "blue") + 
      labs(title = "Seasonally Adjusted Gas Production",
           x = "Time", y = "Seasonally Adjusted Gas Production") +
      theme_minimal()
    ```

e.  Change one observation to be an outlier by running the following snippet:

    ```{r}
    #| eval: TRUE
    # Change to eval: TRUE in order to run
    outlier <- gas |>
      mutate(Gas = if_else(Quarter == yearquarter("2007Q4"), Gas + 300, Gas))
    ```

    Recompute the decomposition. What is the effect of the outlier on the seasonally adjusted data?

    ```{r}
    #| warning: false
    outlier_decomp <- outlier |>
      model(classical_decomposition(Gas, type = "multiplicative"))|>
      components()

     outlier_decomp |>
        autoplot()+
      labs(title = "Classical Decomposition of Gas Production With Outlier",
           x = "Year", y = "Gas Production")+
         theme_minimal()
    ```

    ```{r}
    ggplot(outlier_decomp, aes(x = Quarter, y = season_adjust)) +
      geom_line(color = "blue") + 
      labs(title = "Seasonally Adjusted Gas Production",
           x = "Time", y = "Seasonally Adjusted Gas Production") +
      theme_minimal()
    ```

    Both seasonal and trend pattern change. Also, the seasonally adjust data has a peak in the middle. So the outlier has affected the whole data.

f.  Does it make any difference if the outlier is near the end rather than in the middle of the time series?

```{r}
end_outlier <- gas |>
  mutate(Gas = if_else(Quarter == yearquarter("2010Q2"), Gas + 300, Gas))

outlier2_decomp <- end_outlier |>
  model(classical_decomposition(Gas, type = "multiplicative"))|>
  components()

outlier2_decomp |>
    autoplot()+
  labs(title = "Classical Decomposition of Gas Production With Outlier In The End",
       x = "Year", y = "Gas Production")+
     theme_minimal()
```

```{r}
ggplot(outlier2_decomp, aes(x = Quarter, y = season_adjust)) +
  geom_line(color = "blue") + 
  labs(title = "Seasonally Adjusted Gas Production",
       x = "Time", y = "Seasonally Adjusted Gas Production") +
  theme_minimal()
```

Both seasonal and trend pattern change. Also, the time series seems to be no seasonality.

## Question 7 (STL decomposition, Q3.10 in FPP)

Consider the `canadian_gas` dataset.

a.  Do an STL decomposition of the data.

    ```{r}
    canadian_gas_decomp <- canadian_gas |> 
      model(STL(Volume))|>
      components()
    ```

b.  How does the seasonal shape change over time? \[Hint: Try plotting the seasonal component using `gg_season()`.\]

    ```{r}
    canadian_gas_decomp |>
      autoplot()
    ```

    ```{r}
    #view(canadian_gas_decomp)
    canadian_gas_decomp |>
      gg_season(canadian_gas_decomp$season_year)
    ```

    The time series displays a seasonality. The seasonality changed a bit since 1990. Before that, the volume is higher in Dec and Jan, while it is lower in summer. After 1990, the volume changed dramatically every month.

c.  Apply a calendar adjustment and compute the STL decomposition again. What is the effect on the seasonal shape?

```{r}
canadian_gas_adjusted = canadian_gas |>
  mutate(volume_adj = Volume / days_in_month(Month)) 

canadian_gas_adjusted |>
  model(STL(volume_adj, robust = TRUE)) |>
  components() |>
  autoplot()
```

```{r}
canadian_gas_adjusted |>
  gg_season(y = Volume)
```

The plot is smoother than the previous one. We have ironed out some kinks.
