---
title: "ST5209/X Assignment 3"
format: pdf
editor: visual
author: "Wang Zixuan,A0298286H"
date: "2025-03-21"
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: true
fontsize: 12pt
mainfont: "Times New Roman"
header-includes:
  - \usepackage{titlesec}
  - \usepackage{color}
  - \usepackage{xcolor}
  - \definecolor{DarkRed}{rgb}{0.6,0,0}
  - \definecolor{DarkBlue}{rgb}{0,0,0.6}
  - \titleformat{\section}{\Large\bfseries\color{DarkBlue}}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries\color{DarkBlue}}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\normalsize\bfseries\color{DarkBlue}}{\thesubsubsection}{1em}{}
  - \setlength{\parindent}{0em}
  - \setlength{\parskip}{1em}
geometry: margin=1in
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).
3.  Create a separate folder in the root directory of the repo, label it with your name, e.g. `yanshuo-assignments`
4.  Copy the assignment1.qmd file over to this directory.
5.  Modify the duplicated document with your solutions, writing all R code as code chunks.
6.  When running code, make sure your working directory is set to be the folder with your assignment .qmd file, e.g. `yanshuo-assignments`. This is to ensure that all file paths are valid.[^1]

[^1]: You may view and set the working directory using `getwd()` and `setwd()`.

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(fpp3)
```

## 1. Holt-Winters, residuals, and forecast accuracy

Consider the antidiabetic drug sales time series which can be loaded using the following code snippet.

```{r}
diabetes <- read_rds("../_data/cleaned/diabetes.rds") |>
  select(TotalC)
```

a.  Fit the following exponential smoothing models on the entire time series:
    -   Holt-Winters with multiplicative noise and seasonality,

        for general recipe : \``ETS(X~error(01）+trend(02)+season(03)`\`

        01 is set to be "M" - noise to be multiplicative.

        02 is set to be "A" - linear trend.

        03 is set to be "M" - multiplicative seasonal component

    -   Holt-Winters with a log transformation, with additive noise and seasonality,

        01 is set to be "A" - noise to be additive .

        02 is set to be "A" - linear trend.

        03 is set to be "A" - additive seasonal component

    -   Holt-Winters with multiplicative noise and seasonality, and damping.

        01 is set to be "M" - noise to be multiplicative .

        02 is set to be "Ad" - damped linear trend.

        03 is set to be "M" - multiplicativeseasonal component

```{r}
diabetes <- diabetes |> 
  as_tsibble()

dia_fit <- diabetes |> 
  model(model1 = ETS(TotalC ~ error("M") + trend("A") + season("M")),
        model2 = ETS(log(TotalC) ~ error("A") + trend("A") + season("A")),
        model3 = ETS(TotalC ~ error("M") + trend("Ad") + season("M"))
        )
```

a.  Make ACF plots for the innovation residuals of these three models. Does this suggest that the residuals are drawn from a white noise model?

    ```{r}
    dia_fit |>
      select(model1) |> 
      gg_tsresiduals()

    dia_fit |>
      select(model2) |> 
      gg_tsresiduals()

    dia_fit |>
      select(model3) |> 
      gg_tsresiduals()
    ```

    These residuals do not conform to a white noise model because their ACF exceeds the blue lines and displays a certain periodicity.

b.  Calculate the p-value from a Ljung-Box test on the residuals with lag $h=8$. Does this suggest that the residuals are drawn from a white noise model? What does this mean about the fitted model?

    ```{r}
    dia_fit |>
        augment() |>
        features(.innov, ljung_box, lag = 8)
    ```

    All the p_values are smaller than 0.05. We can reject the null hypothesis. In other words, the models seem to have a bad fit to the data.

c.  Perform time series cross-validation for the three methods, using `.init = 50` and `.step = 10`, and with the forecast horizon $h=4$. Which method has the best RMSSE? How many data points is the error averaged over in total?

    ```{r}
    dia_cv <- diabetes |>
      stretch_tsibble(.step = 10, .init = 50) |>
      model(model1 = ETS(TotalC ~ error("M") + trend("A") + season("M")),
            model2 = ETS(log(TotalC) ~ error("A") + trend("A") + season("A")),
            model3 = ETS(TotalC ~ error("M") + trend("Ad") + season("M"))
            )
            
    dia_fc <- dia_cv |> 
      forecast(h = 4)

    dia_fc |>
      accuracy(diabetes) |>
      select(.model, RMSSE)
    ```

    Holt-Winters with a log transformation, with additive noise and seasonality has the lowest RMSSE.

```{r}
n <- nrow(diabetes)
steps <- ceiling((n - 50) / 10)
steps * 4
```

The error averaged over 64 data points.

## 2. ETS parameters, prediction intervals

The dataset `hh_budget` contains annual indicators of household budgets for a few countries.

a.  Fit `ETS` on the time series comprising savings as a portion of household budgets in Canada. Do not specify what type of ETS model to fit, instead allowing the function to perform automatic model selection.

    ```{r}
    sav_can <- hh_budget |>
      select(Savings) |>
      filter(Country == "Canada")

    sav_fit <- sav_can |>
      model(ETS(Savings))
    ```

b.  Which ETS model was selected?

    ```{r}
    sav_fit
    ```

    ETS model selected Holt-Winters with with additive noise without trend or seasonality automately.

c.  What are the fitted parameters of the model?

    ```{r}
    sav_fit |>
      tidy()
    ```

    The fitted parameters is $\alpha = 0.99989$

d.  Based on the fitted parameters, what simple forecasting method is this similar to?

    The fitted parameters is $\alpha = 0.99989$, which is so close to 1. So this model is similar to the naive method.

e.  Based on your answer to d), how does the width of the prediction interval change as a function of the forecast horizon $h$? Make a plot to verify this.

    ```{r}
    sav_fit |>
      forecast(h = 10) |>
      autoplot()
    ```

The interval width grows with forecatst horizon grows: $\sqrt{h}$.

## 3. Moving averages and differences

Consider the linear trend model $$
X_t = \beta_0 + \beta_1 t + W_t.
$$ Define a time series $(Y_t)$ by taking a moving average of $(X_t)$ with a symmetric window of size 7. Define another times series $(Z_t)$ by taking a difference of $(X_t)$.

a.  What is the mean function for $(Y_t)$? What is the ACVF for $(Y_t)$?
    -   $$
        Y_{t} = \frac{1}{7}\sum_{i=t-3}^{t+3}X_{i}.
        $$

        The mean function for ($Y_{t}$) is

        $$
        E(Y_{t}) =\beta_{0}+\beta_1t.
        $$

        ACVF:

        $$
        \gamma_Y(h)=Cov(Y_t,Y_{t+h}) = \sigma^2\frac{7-|h|}{49},\ \text{for}\ |h|\leq7, 
        $$

        $$
        \gamma_Y(h)=0,\ \text{otherwise}.
        $$
b.  What is the mean function for $(Z_t)$? What is its ACVF?

$$
Z_t=X_t-X_{t-1}.
$$

$$
E(Z_t)=(\beta_0+\beta_1t)-(\beta_0+\beta_1(t-1))=\beta_1.
$$

$$
\gamma_Z(h)=Cov(Z_t,Z_{t+h})=Var(Z_t)=2\sigma^2,\ \text{for}\ h=0; 
$$

$$
\gamma_Z(h)=-\sigma^2,\ \text{for}\ |h|=1;
$$

$$
\gamma_Z(h)=0, \ otherwise  .
$$

a.  What is the CCF of $(Y_t)$ and $(Z_t)$?

$$
\gamma_{YZ}(h)=\frac{\sigma^2}{7}\ \text{for}\ h=3,
$$

$$
\gamma_{YZ}(h)=-\frac{\sigma^2}{7}\ \text{for}\ h=-4,
$$

$$
\gamma_{YZ}(h)=0\ \text{otherwize}.
$$

$$
\gamma_{Y}(0)=\frac{\sigma^2}{7}
$$
$$
\gamma_{Z}(0)=2{\sigma^2}
$$

$$\rho_{YZ}=1/\sqrt14 for h = 3$$

$$\rho_{YZ}=-1/\sqrt14 for h =-4$$
$$
\rho_{YZ}=0\ \text{otherwize}.
$$

a.  Are $(Y_t)$ and $(Z_t)$ jointly stationary?

    Since $(Y_t)$ contains a trend, it is not stationary. $(Z_t)$ is stationary since the process of difference removed the trend. $(Y_t)$ is stationary while $(Z_t)$ is not, they are not jointly stationary.

## 4. Sample vs population ACF

Consider the signal plus noise model $$
X_t = \sin(2\pi t/5) + W_t.
$$

a.  What is the ACF of $(X_t)$?

    $(W_t)$ is white noise, $(W_t)$ and $(W_{t-k})$ $(k\neq0)$ are uncorrelated. Therefore, $(X_t)$ and $(W_t-k)$ $(k\neq0)$ are not correlated.

$$
\rho_x(k)=1\ \text{for} \ k=0;
$$

$$
\rho_x(k)=0\ \text{otherwise}.
$$

a.  Simulate a time series $X_1,X_2,\ldots,X_{200}$ from this model and plot its sample ACF.

    ```{r}
    set.seed(5209)
    n <- 200
    X_t <- tibble(id = 1:n,
                  xt = sin(2 * pi * id/5) + 
                    rnorm(n, mean = 0, sd = 1)) |>
      as_tsibble(index = id)

    X_t |>
      ACF(xt) |>
      autoplot()
    ```

b.  Why does the sample ACF not look like the population ACF function?

    The mean function is not constant, so the time series is not weak stationary. So the sample ACF is different from the population ACF.

c.  Why does the asymptotic normality theorem for the ACF not apply?

    Because the time series is not stationary. The mean function is not constant.

## 5. Gaussian processes

Consider a random vector $(X_1,X_2,X_3,X_4)$ that has the joint Gaussian distribution

$$
\begin{bmatrix} X_1 \\ X_2 \\ X_3 \\ X_4\end{bmatrix} \sim N\left(\begin{bmatrix} -0.5 \\ 0 \\ 0.5 \\ 2\end{bmatrix}, \begin{bmatrix} 3 & 1 & 1 & 1 \\ 1 & 3 & 1 & 1 \\ 1 & 1 & 3 & 2 \\ 1 & 1 & 2 & 3\end{bmatrix} \right)
$$

a.  What is the marginal variance of $X_4$?

    The marginal variance of $X_4$ is 3.

b.  What is the conditional variance of $X_4$, conditioned on the observations $X_1=1,X_2=1,X_3=1$. Does it depend on these particular values? *Hint: The following code snippet creates matrix in R. You may use `solve()` to find the inverse of a matrix.*

    ```{r}
    Sigma <- matrix(c(3, 1, 1, 1,
                      1, 3, 1, 1,
                      1, 1, 3, 2,
                      1, 1, 2, 3), nrow=4, byrow=TRUE)

    Sigma_AA <- Sigma[1:3, 1:3]  # Covariance of X1, X2, X3
    Sigma_AB <- Sigma[1:3, 4]     # Covariance between (X1, X2, X3) and X4
    Sigma_BB <- Sigma[4, 4]       # Variance of X4

    Sigma_AA_inv <- solve(Sigma_AA)

    cond_var_X4 <- Sigma_BB - t(Sigma_AB) %*% Sigma_AA_inv %*% Sigma_AB

    cond_var_X4
    ```

c.  What is the conditional mean of $X_4$, conditioned on the observations $X_1=1,X_2=1,X_3=1$?

    ```{r}
    mu <- c(-0.5, 0, 0.5, 2)
    X_A <- c(1, 1, 1)


    mu_A <- mu[1:3]          # Mean vector of X1, X2, X3
    mu_B <- mu[4]            # Mean of X4
    Sigma_AA <- Sigma[1:3, 1:3]  # Covariance matrix of X1, X2, X3
    Sigma_AB <- Sigma[1:3, 4]     # Covariance vector between (X1, X2, X3) and X4


    cond_mean_X4 <- mu_B + t(Sigma_AB) %*% solve(Sigma_AA) %*% (X_A - mu_A)

    cond_mean_X4
    ```

d.  Write a level 95% prediction interval for $X_4$ given these observations.

```{r}
lower_bound <- cond_mean_X4 - sqrt(cond_var_X4) * 1.96
upper_bound <- cond_mean_X4 + sqrt(cond_var_X4) * 1.96

# Print the result
cat("95% Prediction Interval for X4:", lower_bound, "to", upper_bound, "\n")
```
